Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-03-02 documentation/report.tex
Fixes

I was able to obtain raw data for two dark matter experiments which exhibit these challenges: the PICO-60 bubble chamber \cite{pico}, and the DEAP-3600 liquid argon scintillator \cite{deap}.

@@

In DEAP-3600, photons are absorbed and re-emitted by a wavelength shifting medium in an unpredictable direction, before being detected by one of 255 photomultiplier tubes (PMTs) arranged in a hexagonal lattice on the surface of the spherical detector. The unpredictability severely limits the accuracy of conventional classifiers; the best classifier so far removes 99.6\% of alpha background radiation, while also (undesirably) removing 91.0\% of WIMP events. Because of physical limitations, simulated data is used for calibration, with a limited amount of real-world experimental data available for testing.

@@

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.95\textwidth]{audio_cropped}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.35\textwidth]{image_grid}}}
    \caption{Examples of the audio and image data.}
\end{figure}

@@

In DEAP-3600, the key challenge is the aforementioned unpredictable photon dynamics. The major source of background radiation is ``neck alphas'', which strike atoms in the gas filling the neck above the spherical detector body; because of the unpredictable direction of photon emission, it is non-trivial to calculate the position of these events.

@@

These results confirm that I met my goals on both the PICO-60 and DEAP-3600 experiments.

On PICO-60, the new semi-supervised learning algorithms performed extremely well, exceeding 98\% accuracy while permitting impure calibration data and requiring little manual analysis or optimization.

On DEAP-3600, my cylindrical projection, combined with a 2D CNN, performed better than the conventional classifier on simulated data, reducing the WIMP misclassification rate from 91.0\% to 75.7\% while matching the conventional classifier's 99.6\% background removal rate.

The predictions made on real-world data demonstrate clear potential. Differences between the simulated and real-world data provide potentially useful diagnostic information on the DEAP-3600 simulation (assuming that WIMPs are not present).

>>>>

I was able to obtain raw data for two dark matter experiments which exhibit these challenges: the PICO-60 bubble chamber \cite{pico}, and the DEAP-3600 liquid argon scintillator \cite{deap}. For each experiment, I developed and compared three general-purpose algorithms intended to resolve its inherent challenge (impurity and unpredictable dynamics, respectively).

@@

In DEAP-3600, photons are absorbed by a wavelength shifting medium and re-emitted in an unpredictable direction, before being detected by one of 255 photomultiplier tubes (PMTs) arranged in a hexagonal lattice on the surface of the spherical detector. The randomness severely limits the accuracy of conventional classifiers; the best classifier so far removes 99.6\% of alpha background radiation, while also (undesirably) removing 91.0\% of WIMP events. Because of physical limitations, simulated data is used for calibration, with a limited amount of real-world experimental data available for testing.

@@

@@

In DEAP-3600, the key challenge is the aforementioned unpredictable photon dynamics. The major source of background radiation is ``neck alphas'', which strike atoms in the gas filling the neck above the spherical detector body; because of the unpredictable direction of photon emission, it is non-trivial to identify this class of events.

@@

These results confirm that I met my objective: my new techniques resolved the two fundamental challenges I identified, permitting significantly improved performance, requiring little manual analysis or optimization.

On PICO-60, the new semi-supervised learning algorithms performed extremely well, exceeding 98\% accuracy while nonetheless permitting impure calibration data.

On DEAP-3600, my cylindrical projection, combined with a 2D CNN, performed better than the conventional classifier on simulated data, reducing the WIMP misclassification rate from 91.0\% to 75.7\% while matching the conventional classifier's 99.6\% background removal rate. Additionally, the predictions made on real-world data demonstrate clear potential. Differences between the simulated and real-world data provide potentially useful diagnostic information on the DEAP-3600 simulation (assuming that WIMPs are not present).
